---
title: "ggplot visualizations"
output: html_notebook
---

The benefit of importing data with tidyverse is that you can manipulate your data in your script or Rmd file without changing the underlying raw data.

# setup

```{r pkgs, warning=FALSE}
library(dplyr)
library(ggplot2)
library(readr)
```


# import data

```{r}
# csv with base R

```


```{r message=FALSE}
# tsv with readr
gapminder <- read_tsv("gapminder.tsv", col_names = TRUE)
```


# subset data

#### filter() function

If you're only working with a portion of the data, you want to subset it. Use `filter()` to combine rows. `filter()` is part of the dplyr package in the tidyverse.

```{r}
# working with 2007 data only
gapminder_2007 <- gapminder %>%
    filter(year == 2007)
```

You may also want to summarize data. The `summarize()` function is different from filtering.

#### summarize() function

When subsetting data, you can summarize many rows into a single data point. You can use summarize with the following functions:<br>

* `sum()`
* `mean()`
* `median()`
* `min()`
* `max()`

For example:

```{r}
# average life expectancy in the year 2007 (all countries)
gapminder %>%
    filter(year == 2007) %>%
    summarize(meanLifeExp = mean(lifeExp))
```


You can also summarize multiple variables at once.

```{r}
# 1957 median life expectancy and max GDP per capita
gapminder %>%
    filter(year == 1957) %>%
    summarize(medianLifeExp = median(lifeExp),
              maxGdpPercap = max(gdpPercap))
```


#### group_by() function

# scatterplot

```{r}
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
    geom_point()
```


### log scale axis

```{r}
# x-axis distribution above is too large, add log scale
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    scale_x_log10()
```


# aesthetics

### color

```{r color}
# mapping continent variable to color aesthetic
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp,
                           color = continent)) +
    geom_point() +
    scale_x_log10()
```

### size

```{r size}
# adding size aesthetic to plot above
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp,
                           color = continent,
                           size = pop)) +
    geom_point() +
    scale_x_log10()
```


# faceting - multiple plots

`~` means "by", so facet by ...

```{r}
# facet by continent
# removed color aes, explanation below
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    scale_x_log10() +
    facet_wrap(~ continent)
```

We removed the color aesthetic in the code above because it doesn't add any new information. Here's the same code, but keeping the color aesthetic. You can see that it just creates a useless legend.

```{r}
# facet by continent, but keeping color aesthetic mapped to continent
ggplot(gapminder_2007, aes(x = gdpPercap, y = lifeExp,
                           color = continent)) +
    geom_point() +
    scale_x_log10() +
    facet_wrap(~ continent)
```


Note the change in our data. Because we are faceting by year, we need to use the full dataset and not just a single year's data.

```{r}
# facet by year
# using gapminder data, not gapminder_2007, added color aesthetic back
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp,
                      color = continent)) +
    geom_point() +
    scale_x_log10() +
    facet_wrap(~ year)
```


